\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}

\title{The Global Rank Hinge for Tseitin Resolution}
\author{Inacio F. Vasquez}
\date{}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{question}{Question}

\begin{document}
\maketitle

\section{Context}

Let $G=(V,E)$ be a connected $d$-regular expander and
$\mathrm{TS}(G,f)$ an unsatisfiable Tseitin formula.

Resolution operates over edge variables $\{x_e : e\in E\}$.

For a clause $C$, define its GF(2) incidence vector
$\chi(C)\in\mathbb{F}_2^{E}$ by edge support.

For a live configuration $S_t$ in a resolution derivation define
\[
r_{\mathrm{sim}}(t)
=
\dim_{\mathbb{F}_2}
\mathrm{span}\{\chi(C) : C\in S_t\}.
\]

\section{Closed Direction: Wide Clauses Are Cheap}

\begin{theorem}[Wide Clause Rank Bound]
For any cut $U\subseteq V$,
there exists a polynomial-length resolution derivation of the
cut-parity clause for $U$ such that
\[
\max_t r_{\mathrm{sim}}(t) \le d+1.
\]
\end{theorem}

Therefore, deriving a single clause of width $\Theta(n)$
does not force large simultaneous GF(2) rank.

This direction is closed.

\section{The Global Rank Hinge}

The exponential lower bound for Tseitin resolution
arises only at the level of refuting $\bot$.

The remaining structural question is:

\begin{question}[Global Rank Hinge]
In every resolution refutation of $\mathrm{TS}(G,f)$,
does there exist a time $t$ such that
\[
r_{\mathrm{sim}}(t) = \Omega(n)?
\]
\end{question}

\section{Equivalence to ED* Survival}

Let ED* denote the entropy-depth obstruction
based on simultaneous GF(2) rank.

\begin{proposition}
The ED* program survives for Tseitin formulas
if and only if the Global Rank Hinge question
has an affirmative answer.
\end{proposition}

\begin{proof}
If no such $t$ exists,
then every refutation maintains
$r_{\mathrm{sim}}(t)=O(1)$,
so entropy accumulation fails,
collapsing ED*.

If such a $t$ must exist,
then entropy must reach $\Omega(n)$
during every refutation,
preserving ED*.
\end{proof}

\section{Program Position}

The single-clause direction is resolved.

All remaining obstruction content
is concentrated in the Global Rank Hinge.

No further claims are made.

\end{document}
